---
title: "Assignment8-Coexpression"
output: html_document
---

Goal: We want to understand how plants connect the upstream perception of **environmental signals** (in this case the presence of neighbors) and how this information cascades through the biological network of the organism to affect the downstream outputs of physiological and developmental changes, eg. **gene expression**

upstream input to downstream output

Network intuition example

```{r}
cities <- read.delim("us_cities.txt", row.names=1) # be sure to change the path
cities
```

choose distance within 1500 miles, assigned to be 1  
distance outside 1500 miles, assigned to be 0  
create adjacency matrix  

```{r}
cities_mat <- cities < 1500 # leave original matrix intact
diag(cities_mat) <- 0 # we do not have to fly within each of cities; automatically change matrix to 1/0 form because T/F (1/0)
cities_mat # check out the adjacency matrix
```

exercise 1:
```{r}
rowSums(cities_mat)
colSums(cities_mat)
```

Chicago has the largest number, so it is the most highly connected

Try extend cities range to 2000

```{r}
cities_mat_2000 <- cities < 2000 # leave original matrix intact
diag(cities_mat_2000) <- 0 # we do not have to fly within each of cities; automatically change matrix to 1/0 form because T/F (1/0)
cities_mat_2000 # check out the adjacency matrix
rowSums(cities_mat_2000)
```

yes it changes to Denver. because Denver is in the midwest where is more ajacent to either west coast cities or east coast cities, within 2000 miles.  


Plot network
```{r}
library(igraph) # load package
# make sure to use the 2000 mile distance cutoff 
cities_graph2 <- graph.adjacency(cities_mat_2000, mode = "undirected")
plot.igraph(cities_graph2)
```

exercise 2:  
total # of nodes: 9  
total # of edges: 20 (look at the cities_graph2 file, near the title)
or
```{r}
sum(cities_mat_2000)/2
```

Exercise 3:  
recalculate using cutoff of 2300  
```{r}
cities_mat_2300 <- cities < 2300 # leave original matrix intact
diag(cities_mat_2300) <- 0 # we do not have to fly within each of cities; automatically change matrix to 1/0 form because T/F (1/0)
cities_mat_2300 # check out the adjacency matrix

sum(cities_mat_2300)/2
```

24 edges obtained

----------------
# gene problem
We will define each gene as a node and the edges between the nodes as some value that we can calculate to make an adjacency matrix  

We can use **similarity in gene expression values** to measure the biological distance between genes. Genes that are expressed in more similar patterns should be closer.  

load data
```{r}
library(tidyverse)
# make sure to change the path to where you downloaded this using wget
DE_genes <- read_csv("DEgenes_GxE.csv")
head(DE_genes) #check out the data

# make sure to change the path to where you downloaded this using wget
brass_voom_E <- read_csv("voom_transform_brassica.csv")
brass_voom_E[,-c(38,42,46)] # remove questionable library columns
head(brass_voom_E)

GxE_counts <- DE_genes %>% select(GeneID) %>% left_join(brass_voom_E) #get count data specifically for the GxE genes
head(GxE_counts)

GxE_counts <- GxE_counts %>% column_to_rownames("GeneID") %>% as.matrix(GxE_counts) # some of the downstream steps require a data matrix
head(GxE_counts[,1:6])
```

start with 5 genes subsets
```{r}
GxE_subset <- GxE_counts[1:5,] # first five genes
# create correlation matrix
GxE_cor <- cor(t(GxE_subset)) # gene across tissue, each gene is a variable, use gene to describe each sample
GxE_cor %>% round(3) # round to three decimal points
# set diagonal to 0
diag(GxE_cor) <- 0
GxE_cor
```

Rank the correlation
```{r}
GxE_rank <- apply(GxE_cor,2,function(x) rank(-abs(x)))
GxE_rank
```

Exercise 4:  
“1” in the [“Bra035334”, “Bra033034”] cell means these two genes have the greatest correlation.  
yes they have different value. 
