---
title: "Assignment8-Coexpression"
output: html_document
---

Goal: We want to understand how plants connect the upstream perception of **environmental signals** (in this case the presence of neighbors) and how this information cascades through the biological network of the organism to affect the downstream outputs of physiological and developmental changes, eg. **gene expression**

upstream input to downstream output

Network intuition example

```{r}
cities <- read.delim("us_cities.txt", row.names=1) # be sure to change the path
cities
```

choose distance within 1500 miles, assigned to be 1  
distance outside 1500 miles, assigned to be 0  
create adjacency matrix  

```{r}
cities_mat <- cities < 1500 # leave original matrix intact
diag(cities_mat) <- 0 # we do not have to fly within each of cities; automatically change matrix to 1/0 form because T/F (1/0)
cities_mat # check out the adjacency matrix
```

exercise 1:
```{r}
rowSums(cities_mat)
colSums(cities_mat)
```

Chicago has the largest number, so it is the most highly connected

Try extend cities range to 2000

```{r}
cities_mat_2000 <- cities < 2000 # leave original matrix intact
diag(cities_mat_2000) <- 0 # we do not have to fly within each of cities; automatically change matrix to 1/0 form because T/F (1/0)
cities_mat_2000 # check out the adjacency matrix
rowSums(cities_mat_2000)
```

yes it changes to Denver. because Denver is in the midwest where is more ajacent to either west coast cities or east coast cities, within 2000 miles.  


Plot network
```{r}
library(igraph) # load package
# make sure to use the 2000 mile distance cutoff 
cities_graph2 <- graph.adjacency(cities_mat_2000, mode = "undirected")
plot.igraph(cities_graph2)
```

exercise 2:  
total # of nodes: 9  
total # of edges: 20 (look at the cities_graph2 file, near the title)
or
```{r}
sum(cities_mat_2000)/2
```

Exercise 3:  
recalculate using cutoff of 2300  
```{r}
cities_mat_2300 <- cities < 2300 # leave original matrix intact
diag(cities_mat_2300) <- 0 # we do not have to fly within each of cities; automatically change matrix to 1/0 form because T/F (1/0)
cities_mat_2300 # check out the adjacency matrix

sum(cities_mat_2300)/2
```

24 edges obtained

----------------
# gene problem
We will define each gene as a node and the edges between the nodes as some value that we can calculate to make an adjacency matrix  

We can use **similarity in gene expression values** to measure the biological distance between genes. Genes that are expressed in more similar patterns should be closer.  

load data
```{r}
library(tidyverse)
# make sure to change the path to where you downloaded this using wget
DE_genes <- read_csv("DEgenes_GxE.csv")
head(DE_genes) #check out the data

# make sure to change the path to where you downloaded this using wget
brass_voom_E <- read_csv("voom_transform_brassica.csv")
brass_voom_E[,-c(38,42,46)] # remove questionable library columns
head(brass_voom_E)

GxE_counts <- DE_genes %>% select(GeneID) %>% left_join(brass_voom_E) #get count data specifically for the GxE genes
head(GxE_counts)

GxE_counts <- GxE_counts %>% column_to_rownames("GeneID") %>% as.matrix(GxE_counts) # some of the downstream steps require a data matrix
head(GxE_counts[,1:6])
```

start with 5 genes subsets
```{r}
GxE_subset <- GxE_counts[1:5,] # first five genes
# create correlation matrix
GxE_cor <- cor(t(GxE_subset)) # gene across tissue, each gene is a variable, use gene to describe each sample
GxE_cor %>% round(3) # round to three decimal points
# set diagonal to 0
diag(GxE_cor) <- 0
GxE_cor
```

Rank the correlation
```{r}
# 2 means cols, 1 means rows
GxE_rank <- apply(GxE_cor,2,function(x) rank(-abs(x))) 
# rank for each column from high to low
GxE_rank
```

Exercise 4:  
(A) “1” in the [“Bra035334”, “Bra033034”] cell means these two genes have the greatest correlation.  

(B) yes they have different value. because we are ranking for every column. In each column, the ranking for particular pair can be different depending on the values within that column.

compare pairwise mutual ranks (average ranks)
```{r}
GxE_MR <- sqrt(GxE_rank * t(GxE_rank))
GxE_MR
```

(C) no, their values are the same. because we are doing above is to multiply the different ranking for each given pair to their counterpart different ranking, such that they will agree on the ranking score. then take the average to obtain the average value which is also the same for each pair. 


Define edge: rank less or equal than 2  

(A)
```{r}
genes_adj_MR2 <- GxE_MR <= 2
diag(genes_adj_MR2) <- 0
genes_adj_MR2
```

(B) Bra035334 is connected to Bra033034 and Bra003598


Exercise 6

genes_adj_MR3
```{r}
GxE_cor <- cor(t(GxE_counts))
GxE_cor %>% round(3) # round to three decimal points
# set diagonal to 0
diag(GxE_cor) <- 0
GxE_cor
```

```{r}
# 2 means cols, 1 means rows
GxE_rank <- apply(GxE_cor,2,function(x) rank(-abs(x))) 
# rank for each column from high to low
GxE_MR <- sqrt(GxE_rank * t(GxE_rank))
GxE_MR
```

```{r}
genes_adj_MR3 <- GxE_MR <= 3 # cut off 3
diag(genes_adj_MR3) <- 0
genes_adj_MR3
```

```{r}
genes_adj_MR10 <- GxE_MR <= 10 # cut off 10
diag(genes_adj_MR10) <- 0
genes_adj_MR10
```

(B)
```{r}
sum(genes_adj_MR3)
sum(genes_adj_MR10)
```

Exercise 7 plot networks for MR3 and MR10
```{r}
gene_graphMR3 <- graph.adjacency(genes_adj_MR3, mode = "undirected") #convert adjacency to graph
comps <- clusters(gene_graphMR3)$membership                        #define gene cluster membership
colbar <- rainbow(max(comps)+1)                                   #define colors
V(gene_graphMR3)$color <- colbar[comps+1]                          #assign colors to nodes
plot(gene_graphMR3, layout = layout.fruchterman.reingold, vertex.size = 6, vertex.label = NA)
  
gene_graphMR10 <- graph.adjacency(genes_adj_MR10, mode = "undirected") #convert adjacency to graph
comps <- clusters(gene_graphMR10)$membership                        #define gene cluster membership
colbar <- rainbow(max(comps)+1)                                   #define colors
V(gene_graphMR10)$color <- colbar[comps+1]                          #assign colors to nodes
plot(gene_graphMR10, layout = layout.fruchterman.reingold, vertex.size = 6, vertex.label = NA)
```

Answer: MR10 has more edges such that the network looks more complicated than MR3. color = membership?

Exercise 8 graph density

```{r}
graph.density(gene_graphMR10)
average.path.length(gene_graphMR10)
```


We get the shortest paths between ALL genes in the network and then print the results. We are interested in visualizing the path between Bra033034 (row number 2) and Bra009406 (column number 7). This is where the 2 and 7 arguments come from in get.shortest.paths()

```{r}
gene_graphMR10 <- graph.adjacency(genes_adj_MR10, mode = "undirected")
distMatrix <- shortest.paths(gene_graphMR10, v = V(gene_graphMR10), to = V(gene_graphMR10))
head(distMatrix)[,1:7]

pl <- get.shortest.paths(gene_graphMR10, 2, 7)$vpath[[1]] # pull paths between node 2 and 7

V(gene_graphMR10)[pl]$color <- paste("green")          # define node color
E(gene_graphMR10)$color <- paste("grey")               # define default edge color
E(gene_graphMR10, path = pl)$color <- paste("blue")    # define edge color
E(gene_graphMR10, path = pl)$width <- 10               # define edge width
plot(gene_graphMR10, layout = layout.fruchterman.reingold, vertex.size = 6, vertex.label = NA)
```

Exercise 9

graph density of 1500 miles range
```{r}
cities_mat

gene_graphcities <- graph.adjacency(cities_mat, mode = "undirected") #convert adjacency to graph
comps <- clusters(gene_graphcities)$membership                        #define gene cluster membership
colbar <- rainbow(max(comps)+1)                                   #define colors
V(gene_graphcities)$color <- colbar[comps+1]                          #assign colors to nodes
plot(gene_graphcities, layout = layout.fruchterman.reingold, vertex.size = 6, vertex.label = NA)

graph.density(gene_graphcities)
average.path.length(gene_graphcities)
```

Find shortest btw SEA and DC in 1500 range
```{r}
gene_graphcities <- graph.adjacency(cities_mat, mode = "undirected")
distMatrix <- shortest.paths(gene_graphcities, v = V(gene_graphcities), to = V(gene_graphcities))
head(distMatrix)[,1:7]

pl <- get.shortest.paths(gene_graphcities, 6, 3)$vpath[[1]] # pull paths between node 2 and 7

V(gene_graphcities)[pl]$color <- paste("green")          # define node color
E(gene_graphcities)$color <- paste("grey")               # define default edge color
E(gene_graphcities, path = pl)$color <- paste("blue")    # define edge color
E(gene_graphcities, path = pl)$width <- 10               # define edge width
plot(gene_graphcities, layout = layout.fruchterman.reingold, vertex.size = 6, vertex.label = NA)
```

Find shortest distance btw SEA and DC with 2300 mile range

```{r}
gene_graphcities <- graph.adjacency(cities_mat_2300, mode = "undirected")
distMatrix <- shortest.paths(gene_graphcities, v = V(gene_graphcities), to = V(gene_graphcities))
head(distMatrix)[,1:7]

pl <- get.shortest.paths(gene_graphcities, 6, 3)$vpath[[1]] # pull paths between node 2 and 7

V(gene_graphcities)[pl]$color <- paste("green")          # define node color
E(gene_graphcities)$color <- paste("grey")               # define default edge color
E(gene_graphcities, path = pl)$color <- paste("blue")    # define edge color
E(gene_graphcities, path = pl)$width <- 10               # define edge width
plot(gene_graphcities, layout = layout.fruchterman.reingold, vertex.size = 6, vertex.label = NA)
```

