---
title: "ggplot"
output:
  html_document:
    keep_md: yes
---

```{r, echo=FALSE}
library(tidyverse)
tomato <- read_csv("Tomato.csv")
head(tomato)
```

see the species names:
```{r}
unique(tomato$species)
```

#### Aesthetic
`color`:  
1. plot petleng against leafleng(scatterplot)
2. color the point by treatment, L or H
```{r}
tomato %>% ggplot(mapping = aes(x=petleng, y=leafleng, color = trt)) + 
  geom_point()
```

`shape`: control the shape of plotted points.  
`fill` : control the color fill of the shape  

Create a plot of int3 vs int4 where color indicates __trt__, and shape indicates __who__ measured the plant.  
```{r}
tomato %>% ggplot(aes(x = int3, y = int4, color = trt, shape = who)) +
           geom_point()
```

`size`: control the size of the plotted point  
create a plot of latitude vs longitude where altitude is indicated by the size of the point and species is indicated by color  
```{r}
tomato %>% ggplot(aes(x=lat, y=lon, size = alt, color = species)) + geom_point()
```

#### change plot setting without refering to the cols  
change dots color to skyblue
```{r}
tomato %>% ggplot(mapping = aes(x=petleng,y=leafleng)) +
  geom_point(color="skyblue") # change in the geom function
```

`shape`: The different plotting shapes are numbered from __0 to 25__  
```{r}
tomato %>% ggplot(mapping = aes(x=petleng,y=leafleng)) +
  geom_point(color="skyblue", shape = 2) # triangle
```

#### Geom() control the type of plot that is made.  
* `Geom_smooth()` : add trend lines to your plots  
```{r}
tomato %>% ggplot(aes(x=lon, y = lat)) + 
  geom_smooth() # only add trendlines, no scatter points; if not specified, it will use geom_smooth(method = "gam")
```

* we can add multiple geom to a plot 
example: trendline + scatter plot
```{r}
tomato %>% ggplot(aes(x=lon, y = lat)) + 
  geom_smooth() +
  geom_point()
```

* add linear regress best fit line to the plot:  
```{r}
tomato %>% ggplot(aes(x=lon, y = lat)) + 
  geom_smooth(method="lm") + # linear model
  geom_point()
```

* `geom_histogram()` and `geom_density()`: create histograms, only take x value, y value is calculated for you  
```{r}
tomato %>% ggplot(aes(x=hyp)) +
  geom_histogram(binwidth = 5)
```

`fill` use control the color used to fill the bars  
```{r}
tomato %>% ggplot(aes(x=hyp)) +
  geom_histogram(fill="red")
```

`geom_density`  
```{r}
tomato %>% ggplot(aes(x=hyp)) + 
geom_density()
```

density plot allows comparision between two density:  
```{r}
tomato %>% ggplot(aes(x=hyp, fill=trt)) +
  geom_density(alpha=.5) # alpha value change the color density 0-1, low to high
```

* `geom_boxplot()` and `geom_violin()`: provide quick summaries of different classes of data  
examine hypocotyl length(hyp) of each species. We can map hypocotyl length to the y-axis and species to the x-axis
```{r}
tomato %>% ggplot(aes(x=species, y=hyp)) +
  geom_boxplot()
```

```{r}
tomato %>% ggplot(aes(x=species, y=hyp)) +
  geom_violin()
```

Make a boxplot showing hypocotyl length for the “H” and “L” treatments
```{r}
tomato %>% ggplot(aes(x=trt, y=hyp)) +
  geom_violin() 
```

```{r}
tomato %>% ggplot(aes(x=species, y=int3, fill = trt)) +
  geom_boxplot()
```

* `geom_col(): make a classic __bar chart__, where the height of the bars corresponds to some value in the data  

1. summarize the data  
```{r}
# write a function to calculate sample mean error (SE)
sem <- function(x, na.rm=FALSE) { # by default na.rm = FALSE
  sd(x,na.rm=na.rm)/sqrt(length(na.omit(x)))
}

int3.mean.sem <- tomato %>%
  group_by(species, trt) %>%
  summarize(mean=mean(int3, na.rm=TRUE), sem=sem(int3, na.rm=TRUE))

int3.mean.sem
```

2. Plot the __bar plot__ for int3 mean across species and treatment  
```{r}
int3.mean.sem %>% ggplot(aes(x=species, y = mean, fill=trt)) +
  geom_col(position="dodge") # plot bar bins side by side instead of stacking up
```

3. Add error bar  
```{r}
int3.mean.sem %>% ggplot(aes(x=species, 
                             y = mean, 
                             fill=trt, 
                             ymax=mean+sem, 
                             ymin=mean-sem)) +
  geom_col(position="dodge") +
  geom_errorbar(position = position_dodge(width=0.9), width=.5)
```

Example: Make a bar chart that shows __average leaf length__ for each accession (acs) and trt combination.  
```{r}
# calculate average leaf length for each combination:
leaf.mean.sem <- tomato %>%
  group_by(acs, trt) %>%
  summarize(mean=mean(int3, na.rm=TRUE), sem=sem(int3, na.rm=TRUE))
leaf.mean.sem

# plot 
leaf.mean.sem %>% ggplot(aes(x=acs, 
                              y = mean, 
                              fill = trt,
                              ymax = mean+sem,
                              ymin = mean-sem)) +  
                              geom_col(position="dodge") +
  geom_errorbar(position = position_dodge(width=0.9), width=.5)
```

* `geom_line()`: data appropriate for bar charts also can be plotted using lines  
```{r}
int3.mean.sem %>% ggplot(aes(x=species,
                             y=mean,
                             color=trt,
                             group=trt,
                             shape=trt,
                             ymax=mean+sem,
                             ymin=mean-sem)) +
  geom_line() +
  geom_errorbar(width=.1) +
  geom_point()
```

however, plotting each species’ reaction to the treatment would make more sense:  
```{r}
int3.mean.sem %>% ggplot(aes(x = trt, 
                             y = mean, 
                             color = species, 
                             group = species, 
                             shape = species)) + geom_line() + geom_point()
```

#### Scale() functions  
* we want to fill the bar manually with different colors  
```{r}
int3.mean.sem %>% ggplot(aes(x=species, y = mean, fill=trt, ymax=mean+sem, ymin=mean-sem)) +
  geom_col(position="dodge") +
  geom_errorbar(position = position_dodge(width=0.9), width=.5) +
  scale_fill_manual(values = c("H"="#999999","L"="#E69F00")) # plus this command
```

* a particular useful one to transform x or y to log scale:  
`scale_y_log10()` and `scale_x_log10()`  

#### Facets 
Previous method: the code below separates the data into “H” and “L” treatments before making the density plot.

```{r}
tomato %>% ggplot(aes(x=int3, fill=trt)) +
  geom_density(alpha=.5)
```

Another way to do this: `facet_wrap()` or `facet_grid()`  
`facet_wrap()` separate two densities into two graphs (in a row)  
```{r}
tomato %>% ggplot(aes(x=int3)) +
  geom_density(fill="lightblue") +
  facet_wrap(~ trt) # separate by treatment L and H
```

modify the code above to separate two densities in two rows: 
```{r}
tomato %>% ggplot(aes(x=hyp)) +
  geom_density(fill="papayawhip") +
  facet_wrap(~ trt, nrow = 2) # specify by nrow = , ncol= ,
```

`facet_grid()` can use __two variables__ to facet and uses those variable to specify the grid of rows and columns:
```{r}
# separate variables in int3 
tomato %>% ggplot(aes(x=int3)) +
  geom_histogram(fill="lawngreen") +
  facet_grid(who ~ trt) + # specify which two variables you want to separate in int3, order matters here, 
  ylim(0,40) # change the y scale limit, or xlim()
```

* Titles and labels  
```{r}
tomato %>% ggplot(aes(x=species,y=int3,fill=trt)) +
  geom_boxplot() +
  ggtitle("Tomato") + # add title to the boxplot
  ylab("Internode 3 (mm)") + # change y axis label
  xlab("Tomato Species") # change x axis label
```

* save your plot to another file:  
R will figure out the appropriate file type from the file extension (pdf, png, jpg, tif)
```{r}
ggsave("~/Desktop/Internode3.pdf", height=6, width = 6)
```

